Local Tuning in Peano Curves-based Global Optimization Scheme

### 2016-09-18 2016/YSC/ysc_paper.tex
Update YSC paper

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\sign}{sign}
\newtheorem{theorem}{Theorem}

@@

The one-dimensional characteristical algorithms represent one of the classes of the
optimization algorithms used widely \cite{optHandbook}. A number of schemes of reduction of the
multidimensional problems to one or several one-dimensional ones, i. e. the \textit{dimension
reduction} schemes have been developed \cite{stronginBook}, \cite{butzCurves}.
\par
In recent years, when building the global optimization algorithms, an increased attention
is paid to accounting for the local properties of the objective function. In the
case of the characteristical schemes, this approach allows developing a mixed algorithm
\cite{mixedAlg} as well as using different estimates of Lipschitz constant in different search
domain \cite{sergLocalTuningFirst}. Originally, the latter approach has been proposed for the one-dimensional
problems, also the investigation of this one when using the nested optimization scheme
has been carried out \cite{nestedLocal}. In all these cases, this approach have demonstrated itself to
be an efficient one, having allowed accelerating the convergence of the
optimization methods essentially. In the present paper, a generalization of the method
for building of the local estimates of Lipschitz constants onto the case of Hölder
functions obtained when reducing the multidimensional optimization problems is considered.

@@

%------------------------------------------------------------------------------
\section{Local Adaptive Estimate of Hölder Constant}
%\label{sect:algorithm}
At it is seen from the scheme of the algorithm, regardless to the local properties
of the optimized one-dimensional function, the same values of estimate of Hölder
constant (\ref{step2}) is used to compute the characteristic of all intervals (\ref{step3_1}), (\ref{step3_2}).
In \cite{sergLocalTuning}, it has been proposed to use different values of \(M\) for each interval,
and also the efficiency of this approach in the case of the optimization of the
one-dimensional functions satisfying Lipschitz condition has been shown. In \cite{nestedLocal},
the application of the adaptive estimates of Lipschitz constants in the multidimensional
nested optimization scheme has been considered. For each interval, the local estimate of the
constant is an additive convolution of the ''global'' and ''local'' ones (\(\gamma\) and \(\lambda\), correspondingly):
\begin{displaymath}
  \begin{array}{lr}
    \lambda_i=\max\{H_{i-1},H_i,H_{i+1}\} \\
    H_i=\frac{|z_i-z_{i-1}|}{\Delta_i} \\
    H^k=\max\{H_i:i=2,\dots ,k\} \\
    \gamma_i=H^k\frac{\Delta_i}{\Delta^{max}} \\
    \Delta^{max}=\max\{\Delta_{i}:i=2,\dots ,k\}
  \end{array}
\end{displaymath}
\begin{equation}
\label{additiveConv}
M_i=r\cdot \max\{H_i, \frac{1}{2}(\lambda_i+\gamma_i),\xi\}
\end{equation}
\(\xi\) is chosen small to prevent the case of the function being constant over search interval.
This variant of convolution does not depend on the parameter \(r\), however, the adaptive convolution:
\begin{equation}
\label{additiveAdaptiveConv}
M_i=r\cdot \max\{H_i, \frac{\lambda_i}{r}+\frac{r-1}{r}\gamma_i,\xi\}
\end{equation}
has been considered in \cite{sergLocalTuning} as well.
If it is known a priori that the optimized function has a complex shape with multiple
local minima, then the initial value of \(r\) is specified to be large enough that
results in the dominance of the ''global'' component \(\gamma\) in the adaptive convolution.

%------------------------------------------------------------------------------
\section{Conditions of Convergence}
%\label{sect:algorithm}
In \cite{sergLocalTuning}, a theorem on the convergence of the method in the case if the objective
function is Lipschitz one is given. However, as a rule, such statements are true in
Hölder metrics as well, therefore, presumably, the following theorem is true:

\begin{theorem}
Assume the objective function \(f(x)\) to  satisfy Hölder condition with finite
constant \(H > 0\), and let x be a limit point of \(\{x_k\}\) generated by the algorithm.
Then, the following assertions hold:
\begin{enumerate}
  \item If \(x\in(0;1)\), then convergence to \(x\) is a bilateral one i. e. there
  exist two infinite subsequences of \(\{x_k\}\) converging to \(x\): one from the
  left, the other from the right;
  \item \(f(x_k) \geqslant f(x)\) for all trial points \(x_k, k \geqslant 1\);
  \item If there exists another limit point \(x^* = x\), then \(f(x) = f(x^*)\);
  \item If the function \(f(x)\) has a finite number of local minima in \([0, 1]\),
  then the point \(x\) is a local optimum one;
  \item (Sufficient conditions for convergence to a global minimizer). Let \(x^*\)
  be a global minimizer of \(f(x)\). If there exists an iteration number \(k^*\)
  such that for all \(k > k^*\) the inequality
  \(M_j(k) > H_j(k)\) holds, where \(H_j(k)\) is Hölder constant for the interval
  \([x_{j(k)-1}, x_{j(k)}]\) containing \(x^*\), and \(M_{j(k)}\) is its estimate.
  Then, the set of limit points of the sequence \(\{x_k\}\) coincides with the set
  of global minimizers of the function \(f(x)\).
\end{enumerate}
\end{theorem}

The proof of this theorem requires further theoretical studies. Within the frames of
the present work, it is not performed. The convergence has been established numerically only.

@@

%------------------------------------------------------------------------------
\section{Acknowledgements}
This research was supported by the Russian Science Foundation, project No 16-11-10150
“Novel efficient methods and software tools for the time consuming decision making problems
with using supercomputers of superior performance”.

>>>>

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\sign}{sign}
\newtheorem{hypothesis}{Hypothesis}

@@

The one-dimensional characteristical algorithms represent one of the classes of the
optimization algorithms used widely \cite{optHandbook}. A number of schemes of reduction of the
multidimensional problems to one or several one-dimensional ones, i. e. the \textit{dimension
reduction} schemes have been developed \cite{evolvents2013}, \cite{gergelGrishaginStringin2013}.
All these schemes can be efficiently parallelized \cite{optParallelBook}.
\par
In recent years, when building the global optimization algorithms, an increased attention
is paid to accounting for the local properties of the objective function. In the
case of the characteristical schemes, this approach allows developing a mixed algorithm
\cite{mixedAlg} as well as using different estimates of Lipschitz constant in different search
domain \cite{sergLocalTuningFirst}. Originally, the latter approach has been proposed for the one-dimensional
problems, also the investigation of this one when using the nested optimization scheme
has been carried out \cite{nestedLocal}. In all these cases, this approach have demonstrated itself to
be an efficient one, having allowed accelerating the convergence of the
optimization methods essentially. In the present paper, a generalization of the method
for building of the local estimates of Lipschitz constants onto the case of Hölder
functions obtained when reducing the multidimensional optimization problems is considered.

@@

%------------------------------------------------------------------------------
\section{Local Adaptive Estimate of Hölder Constant}
%\label{sect:algorithm}
At it is seen from the scheme of the algorithm, regardless to the local properties
of the optimized one-dimensional function, the same values of estimate of Hölder
constant (\ref{step2}) is used to compute the characteristic of all intervals (\ref{step3_1}), (\ref{step3_2}).
In \cite{sergLocalTuning}, it has been proposed to use different values of \(M\) for each interval,
and also the efficiency of this approach in the case of the optimization of the
one-dimensional functions satisfying Lipschitz condition has been shown. In \cite{nestedLocal},
the application of the adaptive estimates of Lipschitz constants in the multidimensional
nested optimization scheme has been considered. For each interval, the local estimate of the
constant is an additive convolution of the ``global'' and ``local'' ones (\(\gamma\) and \(\lambda\), correspondingly):
\begin{displaymath}
  \begin{array}{lr}
    \lambda_i=\max\{H_{i-1},H_i,H_{i+1}\} \\
    H_i=\frac{|z_i-z_{i-1}|}{\Delta_i} \\
    H^k=\max\{H_i:i=2,\dots ,k\} \\
    \gamma_i=H^k\frac{\Delta_i}{\Delta^{max}} \\
    \Delta^{max}=\max\{\Delta_{i}:i=2,\dots ,k\}
  \end{array}
\end{displaymath}
\begin{equation}
\label{additiveConv}
M_i=r\cdot \max\{H_i, \frac{1}{2}(\lambda_i+\gamma_i),\xi\}
\end{equation}
\(\xi\) is chosen small to prevent the case of the function being constant over search interval.
This variant of convolution does not depend on the parameter \(r\), however, the adaptive convolution:
\begin{equation}
\label{additiveAdaptiveConv}
M_i=r\cdot \max\{H_i, \frac{\lambda_i}{r}+\frac{r-1}{r}\gamma_i,\xi\}
\end{equation}
has been considered in \cite{sergLocalTuning} as well.
If it is known a priori that the optimized function has a complex shape with multiple
local minima, then the initial value of \(r\) is specified to be large enough that
results in the dominance of the ``global'' component \(\gamma\) in the adaptive convolution.

%------------------------------------------------------------------------------
\section{Hypothesis about Convergence}
%\label{sect:algorithm}
In \cite{sergLocalTuning}, a theorem on the convergence of the method in the case if the objective
function is Lipschitz one is given. However, as a rule, such statements are true in
Hölder metrics as well, therefore, presumably, the following hypothesis is true:

\begin{hypothesis}
Assume the objective function \(f(x)\) to  satisfy Hölder condition with finite
constant \(H > 0\), and let x be a limit point of \(\{x_k\}\) generated by the algorithm.
Then, the following assertions hold:
\begin{enumerate}
  \item If \(x\in(0;1)\), then convergence to \(x\) is a bilateral one i. e. there
  exist two infinite subsequences of \(\{x_k\}\) converging to \(x\): one from the
  left, the other from the right;
  \item \(f(x_k) \geqslant f(x)\) for all trial points \(x_k, k \geqslant 1\);
  \item If there exists another limit point \(x^* = x\), then \(f(x) = f(x^*)\);
  \item If the function \(f(x)\) has a finite number of local minima in \([0, 1]\),
  then the point \(x\) is a local optimum one;
  \item (Sufficient conditions for convergence to a global minimizer). Let \(x^*\)
  be a global minimizer of \(f(x)\). If there exists an iteration number \(k^*\)
  such that for all \(k > k^*\) the inequality
  \(M_j(k) > H_j(k)\) holds, where \(H_j(k)\) is Hölder constant for the interval
  \([x_{j(k)-1}, x_{j(k)}]\) containing \(x^*\), and \(M_{j(k)}\) is its estimate.
  Then, the set of limit points of the sequence \(\{x_k\}\) coincides with the set
  of global minimizers of the function \(f(x)\).
\end{enumerate}
\end{hypothesis}

The proof of this hypothesis requires further theoretical studies. Within the frames of
the present work, it is not performed. The convergence has been established numerically only.

@@

%------------------------------------------------------------------------------
\section{Acknowledgements}
This research was supported by the Russian Science Foundation, project No 16-11-10150
``Novel efficient methods and software tools for the time consuming decision making problems
with using supercomputers of superior performance''.
